---
title: "Report"
output: pdf_document
---

```{r setup, include = FALSE}
library(here)
library(lightgbm)
library(ggplot2)
library(GGally)
source(here::here("data", "data.R"))
source(here::here("src", "trees", "Tree_analysis.R"))

knitr::opts_chunk$set(
  fig.width = 7, fig.height = 4.5, out.width = "92%",
  dpi = 140, message = FALSE, warning = FALSE, echo = FALSE,
  fig.align = "center"
)
```

## Introduction
This project uses the following dataset: https://www.kaggle.com/datasets/udaymalviya/bank-loan-data/data. Each model developed will attempt to predict a borrower's loan status (i.e., wether a loan is fully paid or default) based on financial and demographic predictors, and this is thus a binary classification problem. The goal of the project is to:

- Compare the performance of multiple classification models in terms of predictive accuracy, interpretability and computational cost.
- Find the most important variables for loan_status and compare the demographic and financial predictors.

## Data

### Data wrangling
The dataset was without any NA-s, most importantly there were no NA-s for the loan_status. It did however have some potential logical outliers. We identified and removed observations that satisfied at least one of these conditions:

$
\[
\begin{aligned}
\text{person\_age} &< 18 \\
\text{person\_age} &> 100 \\
\text{person\_income} &< 0 \\
\text{loan\_amnt} &\leq 0 \\
\text{loan\_int\_rate} &\leq 0 \\
\text{loan\_percent\_income} &< 0 \\
\text{person\_emp\_exp} &< 0 \\
\text{person\_emp\_exp} &> \text{person\_age} \\
\text{cb\_person\_cred\_hist\_length} &< 0 \\
\text{cb\_person\_cred\_hist\_length} &> \text{person\_age}
\end{aligned}
\]
$

We transformed the datatype of each column that was supposed to be a factorial to the factorial datatype, as this was processed as a character and integer in R. In addition we added the predictors emp_exp_to_age and cred_hist_to_age, which is employee experience to age ratio and credit history to age ratio respectively. These predictors might also be relevant for the models. 

Lastly we created the train and test split. Since the default to non-default ratio is skewed (prevelance is low as seen in data analysis), we chose to stratify the splits.

### Data analysis
```{r prevelance, fig.width=5, fig.height=3.2, out.width='65%', fig.align='center'}
ggplot(Loan, aes(x = loan_status)) + geom_bar()

prevelance <- mean(Loan$loan_status == 1)

cat("Prevelance: ", prevelance)
```

```{r, numeric_plot, fig.width=7, fig.height=5}
num.cols.num <- names(Loan)[sapply(Loan, is.numeric)]
par(mfrow = c(2, 2))

for (column in num.cols.num) {
  hist(Loan[[column]],
        main = paste("Hist, ", column),
        xlab = column)
  boxplot(Loan[[column]],
       main = paste("Box, ", column),
       xlab = column)
}
``` 

```{r, categorical_plot, fig.width=7, fig.height=5}
num.cols.fac <- names(Loan)[sapply(Loan, is.factor)]
par(mfrow = c(2, 2))

for (column in num.cols.fac) {
  plot(Loan[[column]],
      main = column,
      xlab = column)
}
```


## Models

### Decision trees
```{r}
aggregation_models <- readRDS(here::here("models", "aggregation_models.rds"))
```

#### Single tree
```{r prune, fig.width=11, fig.height=6, out.width='100%'}
prune.loan <- aggregation_models$prune.loan

visualize_tree(model = prune.loan, aggregation = FALSE)
performance.prune.loan <- performance_tree(model = prune.loan, test = test)
```

#### Random forest

```{r rf}
rf_tuning <- readRDS(here::here("models", "rf_tuning.rds"))

plot(rf_tuning$mtry.oob$mtry, rf_tuning$mtry.oob$oob.error, type = "b", xlab = "mtry", ylab = "oob error")
points(rf_tuning$mtry.oob$mtry[rf_tuning$best.mtry.idx], rf_tuning$mtry.oob$oob.error[rf_tuning$best.mtry.idx], col = "red", pch = 20)
legend("topright",
       legend = "Minimum oob error",
       col = "red",
       pch = 20,
       bty = "n")

rf.loan <- aggregation_models$rf.loan

visualize_tree(model = rf.loan, aggregation = TRUE)
performance.rf.loan <- performance_tree(model = rf.loan, test = test)
```
We check positives ratio since this is baseline AUC for PR. The baseline ratio is 
```{r baseline_pr}
cat("Prevelance: ", mean(test$loan_status == 1))
```

#### Boosting

##### Ada Boost
```{r, adaboost}
adaboost <- readRDS(here::here("models", "adaboost.rds"))

ada.loan <- adaboost$ada.model
ada.params <- adaboost$ada.params

print(ada.params)
print(summary(ada.loan))
performance.adaboost.loan <- performance_boost_tree(model = ada.loan, test = test)
```

##### lgbm
```{r, lgbm}
lgbm.data <- readRDS(here::here("models", "lgbm.rds"))
lgbm.model <- lgb.load(here::here("models", "lgbm_model.txt"))

lgbm.params <- lgbm.data$lgbm.params
lgbm.loan <- lgbm.model

print(lgbm.params)
print(summary(lgbm.loan))
performance.lgbm.loan <- performance_boost_tree(model = lgbm.loan, test = test)
```

```{r, bag_lgbm}
bag_lgbm.loan <- readRDS(here::here("models", "bag_lgbm.rds"))

dev.bag_lgbm.loan <- bag_lgbm.loan$prob_dev
cat("Deviation in bootstrap: \n ")
print(summary(dev.bag_lgbm.loan))

performance.bag_lgbm.loan <- performance_boost_tree(bag_lgbm.loan, test)
```

#### Tree summary
```{r}
performance.df <- rbind(
  data.frame(model = "prune", as.data.frame(performance.prune.loan)),
  data.frame(model= "random forest", as.data.frame(performance.rf.loan)),
  data.frame(model = "ada boost", as.data.frame(performance.adaboost.loan)),
  data.frame(model = "lgbm", as.data.frame(performance.lgbm.loan)),
  data.frame(model = "bag lgbm", as.data.frame(performance.bag_lgbm.loan))
)

performance.df.auc_sort <- performance.df[order(-performance.df$auc), ]
cat("AUC: \n")
print(performance.df.auc_sort[, c("model", "auc")], row_names = FALSE)

performance.df.misclass_sort <- performance.df[order(-performance.df$misclass_rate), ]
cat("Misclassification rate: \n")
print(performance.df.misclass_sort[, c("model", "misclass_rate")], row_names = FALSE)

performance.df.sens_sort <- performance.df[order(-performance.df$sensitivity), ]
cat("Sensitivity: \n")
print(performance.df.sens_sort[, c("model", "sensitivity")], row_names = FALSE)

performance.df.spec_sort <- performance.df[order(-performance.df$specificity), ]
cat("Specificity: \n")
print(performance.df.spec_sort[, c("model", "specificity")], row_names = FALSE)
```