---
title: "Report"
output:
  pdf_document: default
author: 
  - Kristoffer Eide
  - Michael Birkeland
date: "April 2026"
header-includes:
  - \usepackage[ruled,vlined,linesnumbered]{algorithm2e}
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include = FALSE}
library(here)
library(lightgbm)
library(ggplot2)
library(GGally)
library(reshape2)
library(class)
library(splines)
source(here::here("data", "data.R"))
source(here::here("src", "trees", "Tree_analysis.R"))
source(here::here("src", "classical", "Classic_performance.R"))

knitr::opts_chunk$set(
  fig.width = 7, fig.height = 3.5, out.width = "92%",
  dpi = 140, message = FALSE, warning = FALSE, echo = FALSE,
  fig.align = "center"
)
```

## Abstract

The purpose of our project is to identify key factors influencing
predictive performance on loan defaulting and to evaluate the predictive
performance of different classification models. This could provide potentially valuable insight, and a good screening system for banks when assessing loan applications. The
predictors include both demographic and financial variables, such as age,
income, credit score, loan type, and interest rate.

The analysis is based of a synthetic dataset containing of one binary
response variable indicating loan status (default or fully paid) and
fifteen predictor variables describing borrower characteristics and loan
attributes, where two of the predictors were added by us. We explored
the data using histograms, boxplots, and barplots to understand variable
distributions and relationships. Additionally, we studied class
imbalance through prevalence and evaluated the relationship between the
predictors using correlation analysis. Multiple classification models
were implemented, including logistical regression, with and without cubic splines, k-nearest neighbors (KNN), decision trees,
random forest, and boosting methods. Model performance was assessed
using metrics such as misclassification rate, sensitivity, specificity,
and area under PR-curves (PR AUC). Model selection and tuning were performed
by implementing K-fold cross-validation. 

The result shows that the tree-based models, outperform the simpler
models in terms of predictive accuracy and AUC. Out of these, the
random forest performed the best for our task.
Key factors of loan default include loan interest rate, previous loan
defaults, loan-to-income ratio, while variables such as credit score,
age and education had a weaker influence. The models indicate
that financial variables related to the loan and the borrower are more
important predictors of default status than demographic characteristics.

## Introduction

This project uses the following dataset:
<https://www.kaggle.com/datasets/udaymalviya/bank-loan-data/data>, which
is synthetic. Each model developed will attempt to predict a borrower's
loan status (i.e., whether a loan is fully paid or default) based on
financial and demographic predictors, and this is thus a binary
classification problem. The goal of the project is to:

-   Compare the performance of multiple classification models in terms
    of predictive accuracy, interpretability and computational cost.
-   Find the most important variables for loan_status and compare the
    demographic and financial predictors.

By answering these questions, it wold hopefully give applicable models for screening loan applicants, and valuable insight in how traits affect loan_status, which could be useful for banks. 

## Data

### Data wrangling

The dataset was without any NA-s, most importantly there were no NA-s
for the loan_status predictor. It did however have some potential logical
outliers. We identified and removed observations that satisfied at least
one of these conditions:
$\texttt{person\_age} < 18$ or $> 100$,
$\texttt{person\_income} < 0$,
$\texttt{loan\_amnt} \le 0$,
$\texttt{loan\_int\_rate} \le 0$,
$\texttt{loan\_percent\_income} < 0$,
$\texttt{person\_emp\_exp} < 0$ or $> \texttt{person\_age}$,
and $\texttt{cb\_person\_cred\_hist\_length} < 0$ or $> \texttt{person\_age}$.

We transformed the datatype of each column that were supposed to be a
categorical to the factor datatype, as this was processed as
character and integer types in R. In addition we added the predictors
emp_exp_to_age and cred_hist_to_age, which is employee experience to age
ratio and credit history to age ratio respectively. These predictors
might also be relevant for the models.

Lastly we created the train and test split. Since the default to
non-default ratio is skewed (prevalence is low as seen in data
analysis), we chose to stratify the splits.

### Data analysis

```{r summary, fig.height = 8}
print(summary(Loan))
```

We find the prevalence of successfully paid loans (1) to defaults (0):

```{r prevalence, fig.width=5, fig.height=2, out.width='65%', fig.align='center'}
ggplot(Loan, aes(x = loan_status)) + geom_bar()

prevalence <- mean(Loan$loan_status == "1")

cat("Prevelance: ", prevalence)
```

The prevalence is quite low, which makes the dataset skewed. Then
misclassification rate will be a less optimal measurement, as a poor
baseline model always predicting the majority class will have a
misclassification rate of 0.22, which is already low.

The ROCR plots the true positive ratio (TPR) against the false positive
ratio (FPR), where \begin{align*}
\text{TPR} &= \frac{\text{TP}}{\text{TP} + \text{FN}} \\
\text{FPR} &= \frac{\text{FP}}{\text{FP} + \text{TN}}
\end{align*}

When the number of negatives is high, thus giving low prevalence, TN
also tends to be high, and FPR can then be relatively low despite of FP being high. A constant low FPR, means the TPR can increase
by a lot without FPR increasing too much when adjusting the decision
thresholds. This gives a high ROCR curve area, and thus high AUC. The AUC
baseline would be 0.5, independent of the prevalence, although low
prevalence affects the ROCR curve optimistically. The ROCR can, in other
words, be too optimistic despite the model producing many false
positives.

We define prevalence and precision as \begin{align*}
\text{prevalence} &= \frac{\text{TP} + \text{FN}}{\text{TP} + \text{FN} + \text{FP} + \text{TN}} \\
\text{precision} &= \frac{\text{TP}}{\text{TP} + \text{FP}}
\end{align*}

The PR curve plots precision to recall (TPR). When positives is rare (TN
is high, which also gives low prevalence), even a moderate number of FP
would severely affect the precision. Thus giving a lower AUC score,
which makes the PR sensitive to low prevalence. The baseline AUC would
in this case be equal to the prevalence.

The strength of the ROCR is that it gives an interpretable result of how
well the model separates the two classes, and it's baseline is not
dependent on the prevalence, making it comparable across datasets. If the prevalence would be relatively even,
we would therefore usually prefer the ROCR measurements. We would actually prefer the same if the
prevalence is very high, as the precision then becomes very high in
spite of the model producing many false positives.

Next we look at the logically most important numeric and categorical
predictors. There might prove to be more important predictors later, but
these are the ones we intuitively think should affect the data the most.

```{r numeric_plot, fig.width=7, fig.height=7}
key_numeric <- c("loan_percent_income", "loan_int_rate", "credit_score")
num.cols.num <- names(Loan)[sapply(Loan, is.numeric)]
par(mfrow = c(3, 3), mar = c(3, 3, 2, 1), oma = c(0, 0, 4, 0), cex = 0.6)

for (column in key_numeric) {
  hist(Loan[[column]], prob = TRUE, border = "black",
       main = paste("Hist, ", column),
       xlab = column)
  lines(density(Loan[[column]]), col = "red", lwd = 2)
  boxplot(Loan[[column]],
       main = paste("Box, ", column),
       xlab = column)
  boxplot(Loan[[column]] ~ Loan$loan_status,
          main = paste("Box, ", column, " by loan_status"),
          xlab = column)
}

mtext("Numeric predictors", side = 3, outer = TRUE, line = 1.2, cex = 1.3)
```

```{r categorical_plot, fig.width=7, fig.height=5}
key_categorical <- c("previous_loan_defaults_on_file", "loan_intent", "person_home_ownership") 
par(mfrow = c(3, 2), oma = c(0, 0, 4, 0), cex = 0.5)

for (column in key_categorical) {
  barplot(table(Loan[[column]]),
      main = paste("Bar, ", column),
      xlab = column)
  rate <- tapply(Loan$loan_status == "1", Loan[[column]], mean)
  barplot(rate, main = paste("Default rate by", column), xlab = "column", ylab = "P(default)")
}

mtext("Categorical predictors", side = 3, outer = TRUE, line = 1.2, cex = 1.3)
```

We see that loan_percent_income is right-skewed from the histogram and
boxplot. These large outliers might have disproportionate influence on
the predicted values or coefficients in KNN or log-regression, but is
less dangerous for the tree based models. For our classic models, we
would therefore either scale the numeric predictors for KNN or remove
outliers or log transform numeric predictors for log-regression. The
distribution for loan_int_rate seems to be multimodal, suggesting there
may be hidden categories affecting the distribution. Therefore, splines
could be a good idea for the classic model, while the tree models are
quite insensitive to this since they are non-parametric. The
credit_score seems pretty unimodal, but we see that unlike
loan_percent_income and loan_int_rate, the boxplots do not vary much by
group, indicating that this predictor might have little effect on the
loan_status.

The previous_loan_defaults_on_file looks suspicious since no one who has
previously defaulted defaults again. This dataset is however synthetic,
which might explain the skewness It could also be explained by there being stronger
terms and conditions for people previously defaulting to get another
loan, which makes them less likely to default twice. The loan_intent
seems to have an effect on default_rate, as well as
person_home_owernship. There are little observations on the categoricals
other and own, which can make estimated effect noisy. A possible
solution would be to merge other and own into a rare group. Category own
is however quite large, so we would prefer not to merge this into rare.
Since there is then really only one rare category, the best option would
perhaps be to remove all observations with person_home_ownership =
other. We choose to keep it, to avoid losing data, since the tree models
are not as affected by this, while we consider removing it for the
classical models (KNN, logreg).

Since the dataset contains categorical predictors, and many of the
numeric predictors do not seem to follow a Gaussian distribution, LDA
and QDA seems to be less natural while KNN, log-regression and the tree
models still are plausible.

```{r corr}
par(mfrow = c(1, 1), oma = c(0, 0, 4, 0))

numeric_cols = sapply(Loan, is.numeric)

cormat <- cor(Loan[, numeric_cols])

get_lower_tri <- function(cormat) {
  cormat[upper.tri(cormat)] <- NA
  return(cormat)
}

get_upper_tri <- function(cormat) {
  cormat[upper.tri(cormat)] <- NA
  return(cormat)
}

lower_tri <- get_lower_tri(cormat)

melted_cormat <- melt(lower_tri, na.rm = TRUE)

ggplot(data = melted_cormat, aes(Var1, Var2, fill = value))+ labs(title =  "Numeric predictor correlation plot") + 
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
   name="Pearson\nCorrelation") +
  theme_minimal()+ 
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 9, hjust = 1))+
 coord_fixed()

```

The correlation plot indicates that some of the predictors are highly
correlated, which is undesirable for interpretability. The model might
give arbitrary weight to either predictor across different fits,
resulting in large standard errors and weak interpretability. The
predictors can still be very good for prediction however, as they might
be engineered predictors (e.g. loan_percent_income) which gives the
models clearer patterns to evaluate on. The prediction is often less
hurt by correlation, as it does not care about a clean split between the
weight of the correlated predictors, only the total contribution to the
predicted value. To reduce the amount of correlation between the predictors we consider subset selection or regularization methods.

## Models

### Classical models

```{r logreg}
logreg.loan <- glm(loan_status ~ ., data=train, family="binomial")
coef(logreg.loan)
performance.logreg.loan <- performance_logreg(model = logreg.loan, test = test, threshold = 0.5)

logreg.spline.loan <- glm(loan_status ~ ns(loan_percent_income, df=4) + ns(loan_int_rate, df=4) + ., data=train, family="binomial"
)
performance.logreg.spline.loan <- performance_logreg(model = logreg.spline.loan, test = test, threshold=0.5)

```

```{r knn}
#auc=rep(NA,30)

# for (k in 1:30) {
#performance.knn.loan <- performance_knn(k=k, train = train, test = test)
#   auc[k] = performance.knn.loan$auc
# }
#auc
```

### Decision trees

```{r read_agg}
aggregation_models <- readRDS(here::here("models", "aggregation_models.rds"))
```

**Pruned single tree**

A single decision tree divides the predictor axis into several areas, on
which we predict by majority vote for each observation in the area we
predict on. We use a recursive binary splitting algorithm, which is
computationally efficient, and can be formulated as follows

```{=latex}
\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{stopping\_rule, training\_data, splitting\_criterion}
\Output{fitted decision tree}
\BlankLine
\While{ stopping rule not met}{
    \For{each possible binary split on predictors in each node}{
      evaluate candidate split using splitting criterion 
    }
    choose split with smallest impurity/loss \;
    split the optimal node
}
\caption{Recursive binary split}
\end{algorithm} 
```

The splitting criterion is RSS for regression, and node impurity for
classification, either Gini index $G = \sum_{k = 1}^K p_k(1-p_k)$ or
cross-entropy $D = - \sum_{k=1}^K p_k \log p_k$. Node impurity is
generally better for classification, than misclassification rate.
Consider as an example two classes with 400 nodes each written as (400,
400). If the model were to choose between the splits (100, 300) and
(300, 100), or (200, 400) and (200, 0) the misclassification rate would be
25% for both. We would however prefer the latter, as this has a pure
class (low node impurity), which is to prefer. To reduce the risk of
overfitting we use a stopping rule, like minimum improvement in
splitting criterion, or minimal observations in node after splitting.

To reduce the risk of overfitting, we prune the trees by minimizing a
cost-complexity function. K-fold cross validation is used to find
the optimal $\alpha$, or equivalently the optimal subtree size. The
algorithm for cost complexity and K-fold cross validation is as follows:

```{=latex}
\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tree, $\alpha$}
\Output{pruned decision tree}
\BlankLine
    minimize $C_\alpha(T) = \alpha * |T| + Q(T)$ for different $\alpha$ \;
    choose tree with lowest $C_\alpha$ using cross-validation
\caption{Pruning}
\end{algorithm} 
```

```{=latex}
\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{model, train\_data, params, K}
\Output{optimal param}
\BlankLine
    divide train into K folds \;
    \For{param\_i in params}{
      \For{fold\_j in folds}{
      train model on all folds except fold\_j \;
      calculate validation error $err_{ij}$ on fold\_j \;
  }
  compute mean validation error $\bar{err}_i = \frac{1}{K} \sum_{j=1}^K err_{ij}$ \;
}
\Return{Param with lowest $\bar{err}_i$}
\caption{K-fold cross-validation}
\end{algorithm}
```

```{r prune, fig.width=8, fig.height=4, out.width='100%'}
prune.loan <- aggregation_models$prune.loan

visualize_tree(model = prune.loan, aggregation = FALSE)
performance.prune.loan <- performance_tree(model = prune.loan, test = test)
```

From the decision tree plot we see that the most important predictors
for the tree model is:

\begin{enumerate}
  \item previous\_loan\_defaults\_on\_file
  \item loan\_percent\_income
  \item loan\_int\_rate, person\_home\_ownership
  \item person\_income
\end{enumerate}

**Random forest**

To reduce the variance of our tree models, we can use bagging. We fit
multiple models on bootstrap aggregated data, then average these models.
The variance of the new model is

\begin{align*}
Var[\bar{X}] = Var[\frac 1 n \sum X_i] = \frac{1}{n^2} Var[X_i] + \sum_{i \neq j} Cov(X_i, X_j)
\end{align*}

Increasing n would reduce the first term, however the tree models would
usually choose the same important predictors each time we fit a new model, causing the
models to be correlated and thus increasing the second term. To reduce
the correlation between each tree, we sample a certain fraction of the
predictors during each split in the binary recursive split algorithm,
only allowing the tree to choose between one of these.For classification
we usually choose $m = \sqrt{p}$ and for regression $p/3$ where p is the
number of predictors. The algorithm can be formulated as follows:

```{=latex}
\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{training\_data, stopping\_rule, splitting\_criterion, num\_predictors, num\_trees}
\Output{fitted random forest}
\BlankLine
  \For{$b=1$ to num\_trees}{
    sample nrow(training\_data) bootstrap samples \;
    fit tree on the bootstrap sample using recursive binary splitting algorithm where each split considers a random subset of num\_predictors predictors \;
  }
  predict by averaging each tree
\caption{Random forest}
\end{algorithm} 
```

When bootstrap aggregating the data one can show that $~ 1/3$ of the
train data is left out when training the model. On average,
observation_i would then not be used by $~1/3 \cdot \text{num\_trees}$ of the
trees. Thus we can use these left out observations in each tree to calculate the error, called out of
bag error, instead of using cross validation.

```{=latex}
\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{training\_data, bagging model}
\Output{oob error}
\BlankLine
  n = nrow(training\_data)
  \For{i=1 to n}{
    find each tree that excluded observation i during training \;
    predict response on observation\_i using these trees \;
    calculate error err\_i
  }
\Return($\frac{1}{n} \sum_{i = 1}^{n} err_i$)
\caption{OOB}
\end{algorithm} 
```

Although we usually would choose $\sqrt p = \sqrt{15} \approx 4$, we
examine the best parameter selection size, mtry, using the oob error:

```{r rf tuning}
rf_tuning <- readRDS(here::here("models", "rf_tuning.rds"))

plot(rf_tuning$mtry.oob$mtry, rf_tuning$mtry.oob$oob.error, type = "b", xlab = "mtry", ylab = "oob error")
points(rf_tuning$mtry.oob$mtry[rf_tuning$best.mtry.idx], rf_tuning$mtry.oob$oob.error[rf_tuning$best.mtry.idx], col = "red", pch = 20)
legend("topright",
       legend = "Minimum oob error",
       col = "red",
       pch = 20,
       bty = "n")
```

The optimal size for this dataset turned out to be 6. For the random
forest model with mtry = 6, we get the following variable importance and
performance:

```{r rf performance}
rf.loan <- aggregation_models$rf.loan

visualize_tree(model = rf.loan, aggregation = TRUE)
performance.rf.loan <- performance_tree(model = rf.loan, test = test)
```

### Boosting

The boosting methods are sequential, unlike bagging and random forest
which builds trees independently and aggregate their predictions. In
boosting, models are added one at a time, where each model attempts to
improve on the previous model's error.

**Ada Boost** 

The Ada boost algorithm can be formulated as follows

```{=latex}
\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{training\_data, num\_trees}
\Output{fitted boosted model}
\BlankLine
  Initialize observation weights $w_i = 1/N, i = 1,2,,.., N$
  Fit classifier $G_m(x)$ on training\_data using weights $w_i$
  \For{m = 1,2..., num\_trees}{
    Compute $\text{err}_\text{m} = \frac{\sum_{i=1}^N w_i I(y_i \neq G_m(x))}{\sum_{i=1}^N w_i}$ \;
    Compute $\alpha_\text{m} = \log{(1-\text{err}_\text{m})/\text{err}_\text{m}}$ \;
    Set $w_i <- w_i*\exp{\alpha_\text{m} \cdot I(y_i \neq G_m(x_i))}, i = 1, 2, .., N$
  }
  \Return{$G(x) = sign[\sum_{m=1}^N \alpha_M G_m(x)]$}
\caption{Ada boost}
\end{algorithm} 
```

We used cross validation on the most important parameters for Ada boost, where we found the most important predictors and appropriate grids using LLM GPT-5.4.  Then we got the following parameters: 

```{r adaboost cv}
adaboost <- readRDS(here::here("models", "adaboost.rds"))

ada.loan <- adaboost$ada.model
ada.params <- adaboost$ada.params

print(ada.params)
```

The performance of this model was:

```{r adaboost performance}
performance.adaboost.loan <- performance_boost_tree(model = ada.loan, test = test)
```

**Bagged lgbm** 

Gradient boosting is more general than the Ada boost,
allowing for optimization for any type of loss function. It can be
formulated as follows:

```{=latex}
\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}
\SetKwInOut{Output}{Output}

\Input{training\_data, number of trees $M$, learning rate $\nu$}
\Output{Boosted tree}

\BlankLine
$N = \mathrm{nrow}(\text{training\_data})$\;
Initialize $f_0 = \arg\min_\gamma \sum_{i=1}^N L(y_i,\gamma)$\;

\For{$m=1$ \KwTo $M$}{
    \For{$i=1$ \KwTo $N$}{
        compute $r_{im} = - \frac{\partial L(y_i, f_{m-1}(x_i))}{\partial f(x_i)}$\;
    }
    fit tree on $r_{im}$ giving terminal regions $R_{jm}$, $j = 1, 2, \ldots, J_m$\;
    \For{$j=1$ \KwTo $J_m$}{
        compute $\gamma_{jm} = \arg\min_\gamma \sum_{x_i \in R_{jm}} L(y_i, f_{m-1}(x_i) + \gamma)$\;
    }
    calculate $f_m = f_{m-1} + \nu \sum_{j=1}^{J_m} \gamma_{jm} I(x \in R_{jm})$\;
}

\Return{$\hat{f}(x) = f_M(x)$\;}
\caption{Gradient boosting}
\end{algorithm}
```

For classification we use the multinomial logistic loss function, which in our case is binomial:

\begin{align*}
p_k(x) &= \frac{e^{f_k(x)}}{\sum_{l=1}^K e^{f_l(x)}} \\
L(y, f(x)) &= - \sum_{k=1}^K I(y = k)\,\log\bigl(p_k(x)\bigr)
\end{align*}

Lgbm additionally uses stochastic boosting, where for each tree, just
like random forest, the model is only allowed to use a random subset of
the predictors to reduce variance. We use cross validation to optimize
the most important parameters, were we used LLM GPT-5.4 to evaluate which parameters that generally is the most important, and appropriate parameter grids for these ones. We then got the following paremeters:

```{r lgbm params}
lgbm.data <- readRDS(here::here("models", "lgbm.rds"))
lgbm.model <- lgb.load(here::here("models", "lgbm_model.txt"))

lgbm.params <- lgbm.data$lgbm.params
lgbm.loan <- lgbm.model

print(lgbm.params)
```

With the performance, and importance plot:

```{r lgbm performance}
performance.lgbm.loan <- performance_boost_tree(model = lgbm.loan, test = test)

lgbm.imp <- lgb.importance(lgbm.loan, percentage = TRUE)
lgb.plot.importance(lgbm.imp, top_n = 10, measure = "Gain")
```

To reduce the variance, we tried to bag 100 lgbm models and got the
following performance:

```{r bag_lgbm}
bag_lgbm.loan <- readRDS(here::here("models", "bag_lgbm.rds"))

dev.bag_lgbm.loan <- bag_lgbm.loan$prob_dev
cat("Deviation in bootstrap:  ")
print(summary(dev.bag_lgbm.loan))

performance.bag_lgbm.loan <- performance_boost_tree(bag_lgbm.loan, test)
```

The bagging did not improve the AUC significantly. A possible
explanation is that the lgbm models were already quite similar, which is
reflected in the low bootstrap deviance.

## Summary

A summary of the performance all the models used, sorted by PR AUC, is found in the
following print:

```{r performance}
performance.df <- rbind(
  data.frame(model = "logreg", as.data.frame(performance.logreg.loan)),
  data.frame(model = "prune", as.data.frame(performance.prune.loan)),
  data.frame(model= "random forest", as.data.frame(performance.rf.loan)),
  data.frame(model = "ada boost", as.data.frame(performance.adaboost.loan)),
  data.frame(model = "lgbm", as.data.frame(performance.lgbm.loan)),
  data.frame(model = "bag lgbm", as.data.frame(performance.bag_lgbm.loan))
)

performance.df.auc_sort <- performance.df[order(-performance.df$auc), ]
cat("AUC: \n")
print(performance.df.auc_sort, row_names = FALSE)
```

Compared to the classical models, such as KNN
and logistic regression, the tree-based methods perform substantially
better in almost all aspects of the performance metrics. This is likey because these models are able to capture complex
nonlinear relationships and interactions between predictors, at the cost of interpretability, which
simpler models fail to do adequately. 

AUC and specificity are the most important performance metrics for our goal, combined with misclassification. This is that from a banking perspective, it would be important to catch potential defaulters (negative outcome), and we thus want the specificity to be high. This could also be achieved by simply assuming everyone defaults, thus giving a poor model, and AUC and miscilassification rate will help evaluate this other aspect of the model. We see that the AUC and misclassification rate is quite high for bag lgbm, lgbm and random forest, but that the random forest performs much better with regards to the specificity. The random forest model is more interpretable than the boosting models, and computationally cheap, which is also an advantage. We would therefore prefer the random forest for this dataset, with the following performance:

```{r best_model}
performance_tree(rf.loan, test, verbose = TRUE)
```

The variable importance plot for random forest, lgbm, and the decision tree shows that financial
variables are the dominant predictors for loan default. Specifically, the
predictor variables such as previous loan defaults, loan-to-income ratio,
interest rate, income and home ownership seemed to be the most
influential to the default status. On the other hand, we see there
are variables with less to none influence on the loan defaulting according to the models. The
demographic variables such as gender and education appear to have little
predictive power. Perhaps surprisingly, variables such as credit score
and loan amount also show little importance. This may be due to the
synthetic nature of the data or because their effects already are
captured by other correlated variables
