---
title: "Report"
output: 
  html_document: default
  pdf_document: default
author: 
  - Kristoffer Eide
  - Michael Birkeland
date: "April 2026"
header-includes:
  - \usepackage[ruled,vlined,linesnumbered]{algorithm2e}
---

```{r setup, include = FALSE}
library(here)
library(lightgbm)
library(ggplot2)
library(GGally)
library(reshape2)
source(here::here("data", "data.R"))
source(here::here("src", "trees", "Tree_analysis.R"))

knitr::opts_chunk$set(
  fig.width = 7, fig.height = 3.5, out.width = "92%",
  dpi = 140, message = FALSE, warning = FALSE, echo = FALSE,
  fig.align = "center"
)
```

## Introduction
This project uses the following dataset: https://www.kaggle.com/datasets/udaymalviya/bank-loan-data/data, which is synthtetic Each model developed will attempt to predict a borrower's loan status (i.e., wether a loan is fully paid or default) based on financial and demographic predictors, and this is thus a binary classification problem. The goal of the project is to:

- Compare the performance of multiple classification models in terms of predictive accuracy, interpretability and computational cost.
- Find the most important variables for loan_status and compare the demographic and financial predictors.

## Data

### Data wrangling
The dataset was without any NA-s, most importantly there were no NA-s for the loan_status. It did however have some potential logical outliers. We identified and removed observations that satisfied at least one of these conditions:

\[
\begin{aligned}
\text{person\_age} &< 18 \\
\text{person\_age} &> 100 \\
\text{person\_income} &< 0 \\
\text{loan\_amnt} &\leq 0 \\
\text{loan\_int\_rate} &\leq 0 \\
\text{loan\_percent\_income} &< 0 \\
\text{person\_emp\_exp} &< 0 \\
\text{person\_emp\_exp} &> \text{person\_age} \\
\text{cb\_person\_cred\_hist\_length} &< 0 \\
\text{cb\_person\_cred\_hist\_length} &> \text{person\_age}
\end{aligned}
\]

We transformed the datatype of each column that was supposed to be a categorical to the categorical datatype, as this was processed as a character and integer in R. In addition we added the predictors emp_exp_to_age and cred_hist_to_age, which is employee experience to age ratio and credit history to age ratio respectively. These predictors might also be relevant for the models. 

Lastly we created the train and test split. Since the default to non-default ratio is skewed (prevalence is low as seen in data analysis), we chose to stratify the splits.

### Data analysis
```{r summary, fig.height = 8}
print(summary(Loan))
```

We find the prevalence of successfully paid loans (1) to defaults (0):
```{r prevalence, fig.width=5, fig.height=2, out.width='65%', fig.align='center'}
ggplot(Loan, aes(x = loan_status)) + geom_bar()

prevalence <- mean(Loan$loan_status == "1")

cat("Prevelance: ", prevalence)
```
The prevalence is quite low, which makes the dataset skewed. Then misclassification rate will be a less optimal measurement, as a poor baseline model always predicting the majority class will have a misclassification rate of 0.22, which is already low. 

The ROCR plots TPR to FPR, where
\begin{align*}
\text{TPR} &= \frac{\text{TP}}{TP + FN} \\
\text{FPR} &= \frac{\text{FP}}{FP + TN}
\end{align*}

When the number of negatives is high (thus giving low prevalence), TN also tends to be high, and FPR can then be relatively low despite the number of FP being high. A constant low FPR, means the TPR can increase by a lot without FPR increasing too much when adjusting the decision thresholds. Giving a high ROCR curve area, and thus high AUC. The AUC baseline would be 0.5, independent of the prevalence, although low prevalence affects the ROCR curve optimistically. The ROCR can in other words be too optimistic despite the model producing many false positives. 

We define prevalence and precision as
\begin{align*}
\text{prevalence} &= \frac{\text{TP} + \text{FN}}{\text{TP} + \text{FN} + \text{FP} + \text{TN}} \\
\text{precision} &= \frac{\text{TP}}{\text{TP} + \text{FP}}
\end{align*}

The PR curve plots precision to recall (TPR). When positives is rare (TN is high, which also gives low prevalence), even a moderate number of FP would severely affect the precision. Thus giving a lower AUC score, which makes the PR sensitive to low prevalence. The baseline AUC would in this case be equal to the prevalence. 

The strength of the ROCR is that it gives an interpretable result of how well the model separates the two classes, and it's baseline is not dependent on the prevalence. If the prevalence would be relatively even, we would therefore usually prefer the ROCR measurements. Same if the prevalence is very high, as the precision then becomes very high in spite of the model producing many false positives. 

Next we look at the logically most important numeric and categorical predictors. There might prove to be more important predictors later, but these are the ones we intuitively think should affect the data the most.

```{r, numeric_plot, fig.width=7, fig.height=7}
key_numeric <- c("loan_percent_income", "loan_int_rate", "credit_score")
num.cols.num <- names(Loan)[sapply(Loan, is.numeric)]
par(mfrow = c(3, 3), mar = c(3, 3, 2, 1), oma = c(0, 0, 4, 0), cex = 0.6)

for (column in key_numeric) {
  hist(Loan[[column]], prob = TRUE, border = "black",
       main = paste("Hist, ", column),
       xlab = column)
  lines(density(Loan[[column]]), col = "red", lwd = 2)
  boxplot(Loan[[column]],
       main = paste("Box, ", column),
       xlab = column)
  boxplot(Loan[[column]] ~ Loan$loan_status,
          main = paste("Box, ", column, " by loan_status"),
          xlab = column)
}

mtext("Numeric predictors", side = 3, outer = TRUE, line = 1.2, cex = 1.3)
``` 

```{r, categorical_plot, fig.width=7, fig.height=5}
key_categorical <- c("previous_loan_defaults_on_file", "loan_intent", "person_home_ownership") 
par(mfrow = c(3, 2), oma = c(0, 0, 4, 0), cex = 0.5)

for (column in key_categorical) {
  barplot(table(Loan[[column]]),
      main = paste("Bar, ", column),
      xlab = column)
  rate <- tapply(Loan$loan_status == "1", Loan[[column]], mean)
  barplot(rate, main = paste("Default rate by", column), xlab = "column", ylab = "P(default)")
}

mtext("Categorical predictors", side = 3, outer = TRUE, line = 1.2, cex = 1.3)
```

We see that loan_percent_income is right skewed from the histogram and boxplot. These large outliers might have disproportionate influence on the predicted values or coefficients in KNN or log-regression, but is less dangerous for the tree based models. For our classic models, we would therefore either scale the numeric predictors for KNN or remove outliers or log transform numeric predictors for log-regression. The distribution for loan_int_rate seems to be multimodal, may suggest there are hidden categories affecting the distribution. Therefore splines could be a good idea for the classic model, while the tree models are quite insensitive to this since they are non-parametric. The credit_score seems pretty unimodal, but we see that unlike loan_percent_income and loan_int_rate, the boxplots do not vary much by group, indicating that this predictor might have little effect on the loan_status. 

The previous_loan_defaults_on_file looks suspicious since no one who has previously defaulted defaults again. This dataset is however synthtetic, so this could be why. It could also be explained by there being stronger terms and conditions for people previously defaulting to get another loan, which makes them less likely to default twice. The loan_intent seems to have an effect on default_rate, as well as person_home_owernship. There is little observations on the categoricals other and own, which can make estimated effect noisy.A possible solution would be to merge other and own into a rare group. Category own is however quite large, so we would prefer not to merge this into rare. Since there is then really only one rare category, the best option would perhaps be to remove all observations with person_home_ownership = other. We choose to keep it, to avoid losing data, since the tree models are not as affected by this, while we consider removing it for the classical models (KNN, logreg).

Since the dataset contains categorical predictors, and many of the numeric predictors do not seem to follow a Gaussian distribution, LDA and QDA seems to be less natural while KNN, log-regression and the tree models still are plausible.

```{r corr}
par(mfrow = c(1, 1), oma = c(0, 0, 4, 0))

numeric_cols = sapply(Loan, is.numeric)

cormat <- cor(Loan[, numeric_cols])

get_lower_tri <- function(cormat) {
  cormat[upper.tri(cormat)] <- NA
  return(cormat)
}

get_upper_tri <- function(cormat) {
  cormat[upper.tri(cormat)] <- NA
  return(cormat)
}

lower_tri <- get_lower_tri(cormat)

melted_cormat <- melt(lower_tri, na.rm = TRUE)

ggplot(data = melted_cormat, aes(Var1, Var2, fill = value))+ labs(title =  "Numeric predictor correlation plot") + 
 geom_tile(color = "white")+
 scale_fill_gradient2(low = "blue", high = "red", mid = "white", 
   midpoint = 0, limit = c(-1,1), space = "Lab", 
   name="Pearson\nCorrelation") +
  theme_minimal()+ 
 theme(axis.text.x = element_text(angle = 45, vjust = 1, 
    size = 9, hjust = 1))+
 coord_fixed()

```

The correlation plot indicates that some of the predictors are highly correlated, which is undesirable for interpretability. The model might give arbitrary weight to either predictor across different fits, resulting in large standord errors and weak interpretability. The predictors can still be very good for prediction however, as they might be engineered predictors (e.g. loan_percent_income) which gives the models clearer patterns to evaluate on. The prediction is often less hurt by correlation, as it does not care about a clean split between the weight of the correlated predictors, only the total contribution to the predicted value. To reduce the amount of correlated predictors we can use subset selection or regularization methods.

## Models

### Decision trees

```{r read_agg}
aggregation_models <- readRDS(here::here("models", "aggregation_models.rds"))
```

**Pruned single tree**

A single decision tree divides the precictor axis into seeral areas, on which we predict by majority vote for each observation in the area we predict on. We use a recursive binary splitting algorithm, which is computationally efficient, and can be formulated as follows

\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{stopping\_rule, training\_data, splitting\_criterion}
\Output{fitted decision tree}
\BlankLine
\While{ stopping rule not met}{
    \For{each possible binary split on predictors in each node}{
      evaluate candidate split using splitting criterion 
    }
    choose split with smallest impurity/loss \;
    split the optimal node
}
\caption{Recursive binary split}
\end{algorithm} 

\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{tree, $\alpha$}
\Output{pruned decision tree}
\BlankLine
    minimize $C_\alpha(T) = \alpha * |T| + Q(T)$ for different alpha \;
    choose tree with lowest $C_\alpha$ using cross-validation
\caption{Pruning}
\end{algorithm} 


```{r prune, fig.width=8, fig.height=4, out.width='100%'}
prune.loan <- aggregation_models$prune.loan

visualize_tree(model = prune.loan, aggregation = FALSE)
performance.prune.loan <- performance_tree(model = prune.loan, test = test)
```

**Random forest**

\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{training\_data, stopping\_rule, splitting\_criterion, num\_predictors, num\_trees}
\Output{fitted random forest}
\BlankLine
  \For{$b=1$ to num\_trees}{
    sample nrow(training\_data) bootstrap samples \;
    fit tree on the bootstrap sample using recursive binary splitting where each split considers a random subset of num\_predictors predictors \;
  }
  predict by averaging each tree
\caption{Random forest}
\end{algorithm} 

```{r rf}
rf_tuning <- readRDS(here::here("models", "rf_tuning.rds"))

plot(rf_tuning$mtry.oob$mtry, rf_tuning$mtry.oob$oob.error, type = "b", xlab = "mtry", ylab = "oob error")
points(rf_tuning$mtry.oob$mtry[rf_tuning$best.mtry.idx], rf_tuning$mtry.oob$oob.error[rf_tuning$best.mtry.idx], col = "red", pch = 20)
legend("topright",
       legend = "Minimum oob error",
       col = "red",
       pch = 20,
       bty = "n")

rf.loan <- aggregation_models$rf.loan

visualize_tree(model = rf.loan, aggregation = TRUE)
performance.rf.loan <- performance_tree(model = rf.loan, test = test)
```

### Boosting

**Ada Boost**

\begin{algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{training\_data, num\_trees}
\Output{fitted boosted model}
\BlankLine
  Initialize observation weights $w_i = 1/N, i = 1,2,,.., N$
  Fit classifier $G_m(x)$ on training\_data using weights $w_i$
  \For{m = 1,2..., num\_trees}{
    Compute $\text{err}_\text{m} = \frac{\sum_{i=1}^N w_i I(y_i \neq G_m(x))}{\sum_{i=1}^N w_i}$ \;
    Compute $\alpha_\text{m} = \log{(1-\text{err}_\text{m})/\text{err}_\text{m}}$ \;
    Set $w_i <- w_i*\exp{\alpha_\text{m} \cdot I(y_i \neq G_m(x_i))}, i = 1, 2, .., N$
  }
  output $G(x) = sign[\sum_{m=1}^N \alpha_M G_m(x)]$
\caption{Ada boost}
\end{algorithm} 

```{r, adaboost}
adaboost <- readRDS(here::here("models", "adaboost.rds"))

ada.loan <- adaboost$ada.model
ada.params <- adaboost$ada.params

print(ada.params)
performance.adaboost.loan <- performance_boost_tree(model = ada.loan, test = test)
```

**Bagged lgbm**

\begin{General boosting algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{training\_data, stopping\_rule, splitting\_criterion, num\_predictors, num\_trees}
\Output{fitted random forest}
\BlankLine
  \For{$b=1$ to num\_trees}{
    sample nrow(training\_data) bootstrap samples \;
    fit tree on the bootstrap sample using recursive binary splitting where each split considers a random subset of num\_predictors predictors \;
  }
  predict by averaging each tree
\caption{gradient boosting}
\end{algorithm} 

\begin{Gradient boosting algorithm}[H]
\DontPrintSemicolon
\SetAlgoLined
\SetKwInOut{Input}{Input}\SetKwInOut{Output}{Output}
\Input{training\_data, stopping\_rule, splitting\_criterion, num\_predictors, num\_trees}
\Output{fitted random forest}
\BlankLine
  \For{$b=1$ to num\_trees}{
    sample nrow(training\_data) bootstrap samples \;
    fit tree on the bootstrap sample using recursive binary splitting where each split considers a random subset of num\_predictors predictors \;
  }
  predict by averaging each tree
\caption{gradient boosting}
\end{algorithm} 

```{r, lgbm}
lgbm.data <- readRDS(here::here("models", "lgbm.rds"))
lgbm.model <- lgb.load(here::here("models", "lgbm_model.txt"))

lgbm.params <- lgbm.data$lgbm.params
lgbm.loan <- lgbm.model

print(lgbm.params)
performance.lgbm.loan <- performance_boost_tree(model = lgbm.loan, test = test)
```

```{r, bag_lgbm}
bag_lgbm.loan <- readRDS(here::here("models", "bag_lgbm.rds"))

dev.bag_lgbm.loan <- bag_lgbm.loan$prob_dev
cat("Deviation in bootstrap:  ")
print(summary(dev.bag_lgbm.loan))

performance.bag_lgbm.loan <- performance_boost_tree(bag_lgbm.loan, test)
```

### Summary
```{r performance}
performance.df <- rbind(
  data.frame(model = "prune", as.data.frame(performance.prune.loan)),
  data.frame(model= "random forest", as.data.frame(performance.rf.loan)),
  data.frame(model = "ada boost", as.data.frame(performance.adaboost.loan)),
  data.frame(model = "lgbm", as.data.frame(performance.lgbm.loan)),
  data.frame(model = "bag lgbm", as.data.frame(performance.bag_lgbm.loan))
)

performance.df.auc_sort <- performance.df[order(-performance.df$auc), ]
cat("AUC: \n")
print(performance.df.auc_sort, row_names = FALSE)
```